The role of learning in the evolution of status signaling: a modeling approach

### 2024-02-06 manuscript/manuscript_2.0.tex
Wrapping up new manuscript version

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
  \AddToHook{env/abstract/before}{\vspace*{-4.5cm}}
  \usepackage{float}
  \usepackage{graphicx}
  \usepackage{unicode-math}
  \usepackage{setspace}\doublespacing
  \usepackage[backref=page]{hyperref}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \floatplacement{figure}{H}
  \newcommand{\beginsupplement}{ \setcounter{table}{0}            
  \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  \providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

@@

In dominance and aggressive interactions, a crucial piece of information
to guide individual actions is the fighting ability of interacting
partners, which is sometimes referred to as Resource Holding Potential
(RHP, Parker 1974) or simply as quality. Responsiveness to the quality
of an opponent is central to communication systems in which individuals
exhibit badges of status, where a signal typically conveys quality
(Johnstone and Norris 1993; Rohwer 1975). Such signals tend to be
arbitrary, however, meaning the signals could potentially be produced by
high- and low-quality individuals alike, opening up the possibility of
dishonest signalling and raising the question of evolutionary stability.
Signals can be evolutionarily stable whenever low-quality individuals
pay a higher fitness cost for carrying the signal such that it is not in
the interest of low-quality individuals to fake quality by producing a
large badge. Signal costs inversely proportional to quality may come
about as a result of the production costs of the signal, in which case
the signal works like a handicap (Botero et al. 2010; Johnstone and
Norris 1993; Grafen 1990; Zahavi 1975). Alternatively, the costs can
come from the aggressive reaction of receivers when the convention of
the communication is broken (Enquist, Ghirlanda, and Hurd 2010; Tibbetts
and Dale 2004). For example in paper wasps, subordinate individuals with
experimentally manipulated dominant-like facial patterns received more
aggression from the dominant (Tibbetts and Dale 2004). In this case,
costs are socially rather than developmentally mediated and are
triggered by the violation of the convention, although how the
convention is established in the first place is unclear.

@@

We model the evolution of signals indicating individual quality in the
context of agonistic interactions. For simplicity, we consider a
population of haploid individuals with non-overlapping generations.
Individuals are born every generation with a level of quality (\(Q_i\),
where \(i\) is subscript of the population vector of size \(N\)) given
by a number between zero and one, which is drawn from a truncated normal
distribution \(N( 0.5,\sigma)\). An individual with quality \(0\) has
the lowest RHP, while an individual with quality \(1\) has the highest.
As they develop, individuals produce a phenotypic signal (badge), the
size of which depends on their quality according to a reaction norm
given by \(B_i=1/(1+e^{\alpha_i-\beta_i Q_i})\); where \(\alpha_i\) and
\(\beta_i\) are individual specific traits that determine the shape of
the reaction norm (Fig. \ref{fig:modelStruc} A). Variation in
\(\alpha_i\) and \(\beta_i\) means individuals can have either
uninformative (flat) or informative (logistic) norms. Informative
reaction norms represent a developmental program where the phenotype of
the individual is determined by the environmental conditions under which
it grows, thus causing a correlation between quality and signal. The
size of the signal is constrained to take values between zero and one.
We assume different values of these traits are given by different
alleles and are inherited from mother to offspring unless, with a small
probability (\(\mu\)), mutation changes the allelic value of the
offspring by an amount drawn from a normal distribution
\(TN(0,\sigma_\mu)\).

After birth, individuals go through a round of viability selection. The
survival probability of an individual is given by
\(s_i=1/(1+e^{-k_1-k_2(Q_i-B_i)})\), where \(k_1\) and \(k_2\) are
parameter values determining the shape of the survival function.
Importantly, if \(k_2=0\) survival probability is independent of
quality, while if \(k_2>0\) lower-quality individuals pay a higher price
for similar-sized badges, fulfilling the core assumption of the handicap
principle (Botero et al. 2010; Grafen 1990; Johnstone and Norris 1993).

@@

The decision of whether to escalate a fight against an interacting
partner can depend on the size of the partner's badge, with the
dependency being determined by the focal's experiences acquired through
a learning process. We implement learning using the actor-critic
approach from reinforcement learning (RL) theory (Sutton and Barto 2018;
Quiñones et al. 2020; Leimar and McNamara 2019; Leimar 2021).
Individuals estimate the reward (pay-off) expected from interacting with
partners of different badge sizes (the critic in RL terminology). After
each interaction they update the estimate of reward proportionally to
the difference between their current estimate and the observed reward
(prediction error \(\delta\)) and to the speed of learning
(\(\mathrm{A}\)). Furthermore, individuals express different
probabilities of retreating/attacking depending on the badge size of
their opponent (the actor in RL). They update the probability of
retreating/attacking depending on whether retreating/attacking leads to
an increase in the reward estimation. Thus, if a focal individual
decides to escalate a fight against an individual with a small badge and
this leads to an increase in the reward estimation, then the focal
individual will increase the probability of escalating fights with
individuals of small badges in the future. Given that badge size is a
real number between 0 and 1, there are infinitely many badge sizes.
Thus, the reward estimation, as well as the probability of
retreating/attacking, must be generalized across different values. To
implement generalization we use the linear function approximation method
based on radial basis functions (Sutton and Barto 2018). Specifically,
we pick \(c\) feature centres, which are evenly spaced values along the
badge size interval ({[}0,1{]}) where the updates are focused. For
simplicity, we keep the location of these feature centres constant and
the same for all individuals; they are stored in vector \(\symbf{b}\) .
Each of these feature centres is associated with a weight for reward
estimation and tendency to play retreat in a given interaction. The
reward estimation and the tendency to retreat are calculated (in every
interaction) as the sum of the weights associated with each feature
centre, weighted by the response triggered by the feature (Fig.
\ref{fig:modelStruc} B, dots represent the feature weights). The
response of each feature centre diminishes as a Gaussian function with
the distance between the feature centre and the badge size of the
partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the reward
estimation \(\hat{R}\) when the focal individual (\(i\)) faces
individual \(j\) is given by,\\
\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_j-b_z|}{2\theta^2})} 
\end{equation}

@@

Individuals interact as the focal individual \(n\) times and the
interaction partner is chosen at random. Thus, the expected number of
interactions for each individual is \(2n\). After the interaction round
, individuals in the population reproduce with a probability
proportional to their total pay-off \(w_i\), which is a sum of the
baseline pay-off (\(w_0\)) and all the pay-offs obtained throughout
their life. Thus, the combination of natural selection and genetic drift
changes the distribution of values in \(\alpha\) and \(\beta\) that
segregates in the population, effectively changing the badges expressed
and the communication system.

@@

\caption{Model of communication in the context of aggressive interactions. In A, the reaction norm determines the badge size, *e.g* the size of a black band on a bird's chest. Red shows an informative reaction norm, whereas blue shows a uninformative reaction norm. B individuals perceiving the signal have a behavioral reaction norm that determines their probability of retreating (black line). The reaction norm arises from generalizing the information from the feature weights (black dots). Feature weights are values associated with a particular badge size and are updated as individuals interact with each other. The generalization of the information associated with each feature weight is represented by the grey line and its axis, which shows that the response triggered by the fourth feature diminishes as the value evaluated is further from the feature center. The learning process moves the feature weights (black dots) up or down depending on whether this leads to an increase in the estimated reward. Panels C and D show the effect of learning on the receiver strategy. Receivers in C face signallers with uninformative reaction norms (blue line in A). The uninformative reaction norm yields a badge size of 0.5, thus the behavioural reaction norm in C is only updated around that value. D Receivers face signallers with informative reaction norms (red line in A). Accordingly, receivers develop a threashold-like reaction norm where the decision to retreat increases with the badge size of the interacting partner. Colour scale in C and D indicates the quality of the individual.}\label{fig:modelStruc}
\end{figure}

@@

\caption{Evolution of the badge size as a handicap mediated by learning. Top panels (A.1 and A.2) show the evolutionary dynamics of the sender code. In the y axis the trait value, in the x axis the generation time in thousands of generations. On the left (A.1), changes in the distribution of values for the intercept of the reaction norm ($\alpha$); on the right (A.2) changes in the distribution of the slope ($\beta$). Darker areas of the background correspond to values with high frequency, while orange lines in both panels show the mean of the distribution. Grey lines show the generation time corresponding to the panels below portraing the sender (B) and receiver code (C). In the bottom panels (C1-4), the learned reaction norms correspond to individuals in the population after the interaction round. Color scale indicates quality just as in Fig. \ref{fig:modelStruc}. The black dashed line shows the average reaction norm of the population. On the first half of the simulation (B.1 and B.2) the badge size evolves to its minimum value. In the second half, the slope takes positive values and reaction norms evolve to produce an honest signal of quality. Correspondingly, individuals learn to react to badge size by increasing the probability of retreating with increasing badge sizes}\label{fig:handicap}
\end{figure}

\hypertarget{badges-as-handicaps}{%
\subsubsection{Badges as Handicaps}\label{badges-as-handicaps}}

When we allowed the badge size to evolve (\(\alpha\) and \(\beta\)
changed subject to natural selection and genetic drift) and the signal
worked as a handicap (\emph{i.e.} the cost of the badge was inversely
proportional to the quality of the individual), the sender code often
evolved to produce an honest signal (Fig. \ref{fig:handicap}). The
evolution of the sender code did not happen immediately after the start
of the evolutionary process. Instead, the evolutionary dynamics of the
reaction norm parameters (Fig. \ref{fig:handicap}, A) appeared to
involve a set of steps. First, the intercept of the reaction norm
(\(\alpha\)) describing the relationship between individual quality and
badge size evolved to higher values, reducing the average badge size in
the populations. This is because reaction norms segregating in the
population are flat at the beginning of the simulation; consequently,
the receiver code does not respond to the badge size and larger badges
are costlier and do not trigger lower attack probabilities. During the
first generations, the slope of the sender reaction norm (\(\beta\))
remains close to the initial value of zero. At around generation 4000,
the slope evolves toward positive values. Positive values in the slope
imply that larger badges correlate with higher quality (Fig.
\ref{fig:handicap} B). Receivers then learn to react to such
correlation, increasing the probability of retreating when facing
individuals with larger badges (Fig. \ref{fig:handicap} C). Hence,
natural selection favours larger values of \(\beta\), eventually leading
to an evolutionary equilibrium in which badge size is an honest signal
of quality mediated by the learned responses of receivers (Fig.
\ref{fig:handicap} B.4 and C.4).

The evolutionary trajectory portrayed in figure \ref{fig:handicap} A is
not the only possible outcome. If the slope of the sender reaction norms
(\(\beta\)) evolves, subject to genetic drift, toward negative values
before badges become handicaps, receivers never learn to react to the
size of badges. Thus, badges are only costly and do not provide
information about the quality of individuals (Fig. \ref{fig:noBadge}) .
Eventually, therefore, badges disappear from the population. In
contrast, when the badge does not work as a handicap but instead is
cost-free, the evolutionary process never leads to the establishment of
an honest signal. Instead, subject to genetic drift, the population
either evolves towards the disappearance of the badge or to the maximum
badge size. In either of these cases reaction norms are flat, so the
badge does not provide any information about quality.

\hypertarget{the-evolution-of-polymorphism-mediated-by-learning}{%
\subsubsection{The evolution of polymorphism mediated by
learning}\label{the-evolution-of-polymorphism-mediated-by-learning}}

The amount of information that agents are able to collect via learning
over their life time can strongly change the outcome of evolutionary
dynamics. In the simulations presented in figure \ref{fig:handicap} and
\ref{fig:noBadge}, individuals learned with high speed
(\(\mathrm{A} = 0.4\)) and interacted repeatedly along their lifetime
(2000 interactions on average). When we reduced the number of
interactions that individuals experienced over their lifetime to 300 on
average, we saw a drastic increase in the phenotypic and genetic
variation present in the population. Populations start monomorphic with
a value of zero for both the intercept and the slope of the reaction
norm, and mutations quickly build up a normal distribution around the
starting value. Within the first 2000 generations, the unimodal
distribution in the intercept (\(\alpha\)) splits into a bimodal one.
Later in evolutionary time, one of the peaks splits further, so at the
end of the evolutionary simulation the distribution of the intercept
(\(\alpha\)) in the population shows three distinct peaks. In the case
of \(\beta\), the peaks in the distribution are not so clear-cut, but
the variance of the distribution increases over time. These changes in
the distribution of the parameters of the sender reaction norm imply
that individuals can generally be classified into three distinct types
(Fig. \ref{fig:branching}). Two types express a flat reaction norm with
extreme values for the badge size, meaning that their badge size is not
informative of their quality, whereas the third type shows intermediate
badge sizes determined by individual quality (Fig. \ref{fig:branching}
B.3-4). In this simulation, the values of the slope in the informative
group are negative, which implies that badge size correlates negatively
with quality. Furthermore, the receiver reaction norms developed through
learning, particularly those of individuals of intermediate quality,
respond to the signal of their sender type by increasing the probability
of retreating from a fight with individuals with smaller badges (Fig.
\ref{fig:branching}).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
  \AddToHook{env/abstract/before}{\vspace*{-4.5cm}}
  \usepackage{float}
  \usepackage{graphicx}
  \usepackage{unicode-math}
  \usepackage{geometry}
  \usepackage{setspace}\doublespacing
  \usepackage[backref=page]{hyperref}
  \linespread{2}
  \usepackage{lineno}
  \linenumbers
  \floatplacement{figure}{H}
  \newcommand{\beginsupplement}{ \setcounter{table}{0}            
  \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
  \providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}
  \geometry{top=1in,left=1in,bottom=1in,right=1in}

@@

In dominance and aggressive interactions, a crucial piece of information
to guide individual actions is the fighting ability of interacting
partners, which is sometimes referred to as Resource Holding Potential
(RHP, Parker 1974) or simply as quality. Responsiveness to the quality
of an opponent is central to communication systems in which individuals
exhibit badges of status, where a signal typically conveys quality
(Johnstone and Norris 1993; Rohwer 1975). Such signals tend to be
arbitrary, however, meaning the signals could potentially be produced by
high- and low-quality individuals alike, opening up the possibility of
dishonest signalling and raising the question of evolutionary stability.
Signals can be evolutionarily stable whenever low-quality individuals
pay a higher fitness cost for carrying the signal such that it is not in
the interest of low-quality individuals to fake quality by producing a
large badge. Signal costs inversely proportional to quality may come
about as a result of the production costs of the signal, in which case
the signal works like a `handicap' (Botero et al. 2010; Johnstone and
Norris 1993; Grafen 1990; Zahavi 1975). Throughout this manuscript, we
use the word `handicap' to refer to the differential costs model by
Grafen (1990); however, note that the accuracy of this terminology has
been questioned (Penn and Számadó 2020). Alternatively, the costs can
come from the aggressive reaction of receivers when the convention of
the communication is broken (Enquist, Ghirlanda, and Hurd 2010; Tibbetts
and Dale 2004). For example, in paper wasps, subordinate individuals
with experimentally manipulated dominant-like facial patterns received
more aggression from the dominant (Tibbetts and Dale 2004). In this
case, costs are socially rather than developmentally mediated and are
triggered by the violation of the convention, although how the
convention is established in the first place is unclear.

@@

We model the evolution of signals indicating individual quality in the
context of agonistic interactions. For simplicity, we consider a
population of haploid individuals with non-overlapping generations of
size \(N=3000\). Individuals are born every generation with a level of
quality \(Q_i\) (where \(i\) is subscript of the population vector)
given by a number between zero and one, which is drawn from a truncated
normal distribution \(TN( 0.5,\sigma)\). An individual with quality
\(0\) has the lowest RHP, while an individual with quality \(1\) has the
highest. As they develop, individuals produce a phenotypic signal
(badge), the size of which depends on their quality according to a
logistic reaction norm given by \begin{equation}
B_i=1/(1+e^{-\alpha_i-\beta_i Q_i}) \textrm{;}\label{eq:reacNorm}
\end{equation}

where \(\alpha_i\) and \(\beta_i\) are individual specific traits that
determine the shape of the reaction norm (Fig. \ref{fig:modelStruc} A).
Variation in \(\alpha_i\) and \(\beta_i\) means individuals can have
either uninformative (flat) or informative (logistic) norms. Informative
reaction norms represent a developmental program where the phenotype of
the individual is determined by the environmental conditions under which
it grows, thus causing a correlation between quality and signal. The
size of the signal is constrained to take values between zero and one.
We assume different values of these traits are given by different
alleles and are inherited from mother to offspring unless, with a small
probability (\(\mu\)), mutation changes the allelic value of the
offspring by an amount drawn from a normal distribution
\(N(0,\sigma_\mu)\).

After birth, individuals go through a round of viability selection. The
survival probability of an individual is given by

\begin{equation}
s_i=1/(1+e^{-k_1-k_2(Q_i-B_i)})\textrm{,} \label{eq:viability}
\end{equation}

where \(k_1\) and \(k_2\) are parameter values determining the shape of
the survival function. Importantly, if \(k_2=0\) survival probability is
independent of quality, while if \(k_2>0\) lower-quality individuals pay
a higher price for similar-sized badges, fulfilling the core assumption
of the handicap principle (Botero et al. 2010; Grafen 1990; Johnstone
and Norris 1993).

@@

The decision of whether to escalate a fight against an interacting
partner depends on the size of the partner's badge, with the dependency
being determined by the focal's experiences acquired through a learning
process. Individuals use previous experiences to estimate the reward
expected from encountering a partner with a particular badge size.
However, given that badge size is a real number between 0 and 1,
previous experiences must be generalized across the badge size range. To
implement generalization we use the linear function approximation method
based on radial basis functions (Sutton and Barto 2018). Specifically,
we pick \(c\) feature centres, which are evenly spaced values along the
badge size interval ({[}0,1{]}) where the updates are focused. For
simplicity, we keep the location of these feature centres constant and
the same for all individuals; they are stored in vector \(\symbf{b}\).
Each of these feature centres is associated with a weight for reward
estimation (\(x_z\), \(z\) is an index that goes from 1 to \(c\)) and
tendency to play retreat in a given interaction (\(y_z\)). For each
badge size ecountered the reward estimate and the associated tendency to
retreat are calculated as the sum of the weights associated with each
feature centre, weighted by the response triggered by the feature (Fig.
\ref{fig:modelStruc} B, dots represent the feature weights). The
response of each feature centre diminishes as a Gaussian function with
the distance between the feature centre and the badge size of the
partner (Fig. \ref{fig:modelStruc} B, grey line). Formally, the reward
estimation \(\hat{R}\) when the focal individual (\(i\)) faces
individual \(j\) is given by,

\begin{equation}
\hat{R}_{ij} =  \sum_{z=1}^{c} x_z e^{(-\frac{|B_j-b_z|}{2\theta^2})} 
\end{equation}

@@

The weights vectors for reward estimation (\(\symbf{x}\)) and tendency
to retreat (\(\symbf{y}\)) are the focus of the learning update. In the
baseline implementation of the model, these two vectors are initialized
with values of zero. For the reward estimation (\(\symbf{x}\)), this
means that before any interaction individuals assume they will get a
payoff of zero. As for the behavioral policy, these initial values
translate into a \(0.5\) probability of playing dove. Beyond the
baseline scenario, we explore initializing the policy vector with a)
more positive values triggering more peaceful behaviour, b) more
negative values triggering more aggressive behaviour, c) values that
trigger the ESS of the original \emph{hawk-dove} game and d) letting
these values be defined by evolutionary processes (see Results for more
details). The later was implemented by letting an extra locus pass on
the initial value from parent to offspring and changed by mutation.
Regardless of the initialization, all these values change with every
interaction as a function of the reward experienced. Formally the change
is given by,

\begin{equation}
  \Delta x_z = \Alpha \delta e^{(-\frac{|B_j-b_z|}{2\theta^2})} \textrm{.}
  \label{eq:critic}
\end{equation}

Where \(\delta\) is the \emph{prediction error}, which is the difference
between the predicted and experienced reward, formally

\begin{equation}
  \delta = R - \hat{R} \textrm{.} \label{eq:predError}
\end{equation}

The experienced reward (\(R\)) is the outcome from the \emph{hawk-dove}
game described earlier.

Thus, the learning update depends on the speed of learning (\(A\)) and
the prediction error. The exponential term assures, just as in the
reward estimation, that the size of update is weighted by the distance
of the partner's badge size to each feature centre.

The weights for the probability of retreat (\(\symbf{z}\)) are updated
in a similar fashion according to

\begin{equation}
  \Delta y_z = \Alpha \delta (1-\pi) e^{(-\frac{|B_j-b_z|}{2\theta^2})} \textrm{.}
  \label{eq:actor}
\end{equation}

where \(\pi\) is the probability of choosing the action chosen by the
individual. If an individual chooses to behave as a dove and avoid an
overt conflict,

\begin{equation}
  \pi = q_{ij} = \frac{1}{1+e^{\sum_{z=1}^{c} y_z e^{-\frac{|B_i-b_z|}{2\theta^2}}}} \textrm{;}\label{eq:pi}
\end{equation}

if alternatively, the individual chooses to escalate the fight, \(\pi\)
corresponds to \(1-q_{ij}\). The combination of \(\pi\) and the
\emph{prediction error} (\(\delta\)) assures that behavioral policy
updates are carried increasing the probability of behaviours that
predict higher reward.

This implementation corresponds to the \emph{actor-critic} approach from
reinforcement learning (RL) theory (Sutton and Barto 2018; Quiñones et
al. 2020; Leimar and McNamara 2019; Leimar 2021). In the
\emph{actor-critic} approach the agent updates its behavioral policy
(\emph{actor}) and in parallel its reward estimation (\emph{critic}). In
our model the actor update is implemented by equations \ref{eq:actor}
and \ref{eq:pi}; while the \emph{critic} update is implemented by
equation \ref{eq:critic}. Both of these updates include the
\emph{prediction error} (equation \ref{eq:predError}). The
\emph{prediction error} corresponds to the Rescorla-Wagner rule
(Rescorla and Wagner 1972), which is a key component of learning theory
and widely applied in neurobiology to understand learning processes
(Holland and Schiffino 2016).

Individuals interact as the focal individual \(n\) times and the
interaction partner is chosen at random. Thus, the expected number of
interactions for each individual is \(2n\). After the interaction round,
individuals in the population reproduce with a probability proportional
to their total pay-off \(w_i\), which is a sum of the baseline pay-off
(\(w_0\)) and all the pay-offs obtained from interactions throughout
their life. Thus, the total fitness is composed of the viability
component (equation \ref{eq:viability}) and the total payoff collected
in the interaction round. Therefore, the combination of natural
selection and genetic drift changes the distribution of values in
\(\alpha\) and \(\beta\) that segregates in the population, effectively
changing the badges expressed and the communication system.

@@

\caption{Model of communication in the context of aggressive interactions. In A, the reaction norm  (equation \ref{eq:reacNorm}) determines the badge size, \textit{e.g} the size of a black band on a bird's chest. Red shows an informative reaction norm, whereas blue shows a uninformative reaction norm. B individuals perceiving the signal have a behavioral reaction norm that determines their probability of retreating (black line). The reaction norm arises from generalizing the information from the policy feature weights (black dots). Feature weights are values associated with a particular badge size and are updated as individuals interact with each other. The generalization of the information associated with each feature weight is represented by the grey line and its axis, which shows that the response triggered by the fourth feature diminishes as the value evaluated is further from the feature center. The learning process moves the feature weights (black dots) up or down depending on whether this leads to an increase in the estimated reward. Panels C and D show the effect of learning on the receiver strategy. Receivers in C face signallers with uninformative reaction norms (blue line in A). The uninformative reaction norm yields a badge size of 0.5, thus the behavioural reaction norm in C is only updated around that value. D Receivers face signallers with informative reaction norms (red line in A). Accordingly, receivers develop a threashold-like reaction norm where the decision to retreat increases with the badge size of the interacting partner. Colour scale in C and D indicates the quality of the individual.}\label{fig:modelStruc}
\end{figure}

@@

\caption{Evolution of the badge size as a handicap mediated by learning. Top panels (A.1 and A.2) show the evolutionary dynamics of the sender code. In y axis shows the trait value, in the x axis shows the generation time in thousands of generations. On the left (A.1), changes in the distribution of values for the intercept of the reaction norm ($\alpha$); on the right (A.2) changes in the distribution of the slope ($\beta$). Darker areas of the background correspond to values with high frequency, while orange lines in both panels show the mean of the distribution. Grey lines show the generation time corresponding to the panels below portraing the sender (B) and receiver code (C). In the bottom panels (C1-4), the learned reaction norms correspond to individuals in the population after the interaction round. Color scale indicates quality just as in Fig. \ref{fig:modelStruc}. The black dashed line shows the average reaction norm of the population. On the first half of the simulation (B.1 and B.2) the badge size evolves to its minimum value. In the second half, the slope takes positive values and reaction norms evolve to produce an honest signal of quality. Correspondingly, individuals learn to react to badge size by increasing the probability of retreating with increasing badge sizes.}\label{fig:handicap}
\end{figure}

\hypertarget{badges-as-handicaps-costly-signals}{%
\subsubsection{Badges as Handicaps (costly
signals)}\label{badges-as-handicaps-costly-signals}}

When we allowed the badge size to evolve (\(\alpha\) and \(\beta\)
changed subject to natural selection and genetic drift) and the signal
worked as a handicap (\emph{i.e.} the cost of the badge was inversely
proportional to the quality of the individual), the sender code often
evolved to produce an honest signal (Fig. \ref{fig:handicap}). The
evolution of the sender code did not happen immediately after the start
of the evolutionary process. Instead, the evolutionary dynamics of the
reaction norm parameters (Fig. \ref{fig:handicap}, A) appeared to
involve a set of steps. First, the intercept of the logistic reaction
norm (\(\alpha\)) describing the relationship between individual quality
and badge size evolved to lower values, reducing the average badge size
in the populations. This is because reaction norms segregating in the
population are flat at the beginning of the simulation; consequently,
the receiver code does not respond to the badge size and larger badges
are costlier and do not trigger lower attack probabilities. During the
first generations, the slope of the sender reaction norm (\(\beta\))
remains close to the initial value of zero. At around generation 4000,
the slope evolves toward positive values. Positive values in the slope
imply that larger badges correlate with higher quality (Fig.
\ref{fig:handicap} B). Receivers then learn to react to such
correlation, increasing the probability of retreating when facing
individuals with larger badges (Fig. \ref{fig:handicap} C). Hence,
natural selection favours larger values of \(\beta\), eventually leading
to an evolutionary equilibrium in which badge size is an honest signal
of quality mediated by the learned responses of receivers (Fig.
\ref{fig:handicap} B.4 and C.4).

The evolutionary trajectory portrayed in figure \ref{fig:handicap} A is
not the only possible outcome. If the slope of the sender reaction norms
(\(\beta\)) evolves, subject to genetic drift, toward negative values
before badges become handicaps, receivers never learn to react to the
size of badges. Thus, badges are only costly and do not provide
information about the quality of individuals (Fig. \ref{fig:noBadge}) .
Eventually, therefore, badges disappear from the population. In
contrast, when the badge does not work as a handicap but instead is
cost-free, the evolutionary process never leads to the establishment of
an honest signal. Instead, subject to genetic drift, the population
either evolves towards the disappearance of the badge or to the maximum
badge size. In either of these cases reaction norms are flat, so the
badge does not provide any information about quality. This means that
learning is not sufficient to yield a communication system, but rather
is requires the badge to have a cost proportional to the quality of the
individual.

We have described here two replicate simulations representative of the
two outcomes found in a series of 20 replicates run with the same
parameter combination. The exact evolutionary trajectory differs in each
of these replicates due to the stochastic effects of the dynamical
system. Particularly the timing of the changes was different in each
replicate. Nevertheless, all of the replicates could be classified as
either evolving a functioning communication system, or not having the
badge present in the population.

\hypertarget{the-evolution-of-polymorphism-mediated-by-learning-when-the-signal-is-cost-free}{%
\subsubsection{The evolution of polymorphism mediated by learning when
the signal is
cost-free}\label{the-evolution-of-polymorphism-mediated-by-learning-when-the-signal-is-cost-free}}

The amount of information that agents are able to collect via learning
over their life time can strongly change the outcome of evolutionary
dynamics. In the simulations presented in figure \ref{fig:handicap} and
\ref{fig:noBadge}, individuals learned with high speed
(\(\mathrm{A} = 0.4\)) and interacted repeatedly along their lifetime
(2000 interactions on average). When we reduced the number of
interactions that individuals experienced over their lifetime to 300 on
average, we saw a drastic increase in the phenotypic and genetic
variation present in the population. Populations start monomorphic with
a value of zero for both the intercept and the slope of the reaction
norm, and mutations quickly build up a normal distribution around the
starting value. Within the first 2000 generations, the unimodal
distribution in the logistic intercept (\(\alpha\)) splits into a
bimodal one. Later in evolutionary time, one of the peaks splits
further, so at the end of the evolutionary simulation the distribution
of the logistic intercept (\(\alpha\)) in the population shows three
distinct peaks. In the case of \(\beta\), the peaks in the distribution
are not so clear-cut, but the variance of the distribution increases
over time. These changes in the distribution of the parameters of the
sender reaction norm imply that individuals can generally be classified
into three distinct types (Fig. \ref{fig:branching}). Two types express
a flat reaction norm with extreme values for the badge size, meaning
that their badge size is not informative of their quality, whereas the
third type shows intermediate badge sizes determined by individual
quality (Fig. \ref{fig:branching} B.3-4).
